\setcounter{page}{1}

This thesis provides a general overview of the research that I have been developing since the beginning of my PhD studies in September, 2021. I could define myself as a curious, creative and open-minded person, following the so called \textit{IFISC attitude}, which means that I am always willing to learn new methods and address new problems, even though they are not directly related to my field of expertise. That is why, through this thesis many topics will be covered, from the study of human behavior and social systems, to the study of complex systems and network theory........

\section{\label{sec:scie_lands} Scientific Landscape}

This thesis adress the study of human behavior and social systems from a \textit{complex systems} perspective, which studies the emergence of collective phenomena that arise from the interactions of many individuals, and that cannot be understood by studying the behavior of individual agents in isolation (the so-called \textit{reductionist} approach) \cite{anderson1972more}. The study of collective phenomena has a long history in the natural sciences, specially in the branch of statistical physics \cite{stanley1971phase}. This branch traditionally studies the emergence of collective phenomena in physical systems, such as the phase transitions in magnetic materials via spin models \cite{onsager-1944}, the turbulence in fluids \cite{frisch1995turbulence}, the synchronization in oscillatory systems \cite{pikovsky2001universal}, or percolation \cite{stauffer-1985}. However, in recent years, the study of complex systems has evolved into the study of emergent phenomena beyond physical systems, such as biological \cite{prigogine1977self}, ecological \cite{may-2001}, economic \cite{arthur-1994}, and social systems \cite{castellano-2009}. Even though this branch of physics is relatively young, in 2021, the Nobel Prize in Physics was awarded to Syukuro Manabe, Klaus Hasselmann, and Giorgio Parisi for their contributions to the study of complex systems \cite{nobel-2021}, giving a recognition to this field in the scientific community.

From the migration of birds \cite{roche-1997} to the spreading of a fake news through social media \cite{vosoughi-2018}, there are many examples of collective phenomena beyond the real of physics at which the study of complex systems can be applied. The cascade of failures in power grids \cite{dobson-2007}, the spread of a disease in a population \cite{anderson1991infectious}, the consensus in political elections \cite{anderson-2000}, the emergence of social norms \cite{ellickson-1999} are some examples of social collective behavior in which the global phenomena cannot be understood by looking at a single individual. Social and economic collective phenomena has been studied from a variety of perspectives (sociology, psychology, economics, political sciences...), which often relies on qualitative methods, such as interviews, surveys, or ethnographic studies \cite{bryman-2010}. However, the study of social systems from a complex systems perspective aims to provide a quantitative framework to understand the collective behavior, based on methodologies from statistical mechanics and network theory \cite{newman-book,barabasi-2013}. Nevertheless, this approach needs for a large amount of detailed data to validate theories and develop models, which historically has been a limitation for the study of social systems. It is in fact surprising how other branches of physics, where the typical scale of the phenomena is very large, as astrophysics, or very small, as particle physics, do not suffer from a lack of data, while the study of social systems, where the typical scale is human, has been historically limited by the lack of data.

Thankfully, the digital revolution has changed this picture, allowing the storage of large amounts of data generated by human activities, such as social media, mobile phones, or online platforms. Nowadays, every two year, more human socio-economic data is produced than during all the preceding years of human history together \cite{karsai2019computational}. This data, often referred to as \textit{Big data}, has opened a new era for the study of social systems at a large scale, together with a paradigm shift in the way we understand human behavior \cite{manyika-2011}. Nevertheless, this new era comes with an awareness, as the use of big data for the study of human behavior raises important ethical and privacy concerns, which need to be addressed in order to ensure the responsible use of data for the study of social systems \cite{boyd-2012}. Moreover, from the technical point of view, this huge amount of data needs for a set of computational and mathematical resources to be analyzed and modeled.From this demand, the field of \textit{Computational Social Science} has emerged, with the aim to develop new methods to study human behavior \cite{lazer-2009}. This branch of the complex systems science was born as a combination of methodologies borrowed from social sciences, such as sociology, psychology, or economics, with computational methods from computer science, such as machine learning, data mining, or network theory \cite{watts-2007}. This interdisciplinary approach has allowed to develop new methods for forecasting social phenomena and understanding the basic mechanism behind human interactions. 

One can differentiate two main approaches to build a representation of the reality from the data source. The first one is to focus on the prediction and forecasting of a certain social phenomena, such as the spread of a disease or the price of a stock. In this approach, the data is seen as a necessary input to our methodology to make quantitative predictions \cite{lazer-2009}. However, in this approach, the mechanisms behind the phenomena are often hidden in the data, and the model is seen as a black box that provides accurate predictions \cite{rudin-2019}. In this context, the use of machine learning \cite{murphy-2012} and deep learning \cite{goodfellow-2016} models are  very popular, as they are able to capture complex patterns in the data. The second approach is to focus on the understanding of the mechanisms behind the phenomena. In this approach, the data is seen as a problem to be understood, an observation from which we can extract qualitative behaviors and patterns \cite{axelrod-1997}. In this context, the aim is to develop very simple models that are able to reproduce the main features of the data, and to extract the basic mechanisms behind the phenomena.

Following the later approach, network science has a critical role in the study of socio-economic systems, as it provides a natural framework to study the interactions between individuals. A network, or graph, is a mathematical representation of a set of nodes (individuals) connected by links (interactions), which allows to study the structure of the interactions and the dynamics of the system. The study of networks has a long history in the natural sciences, from the neurons network in the brain \cite{sporns-2004} to food webs in an ecosystem \cite{ings-2008, elith-2009, bastolla-2009}. However, in recent years, new data sources lead to the discovery that complex properties and heterogeneities, present in most social systems, need for a topological description in terms of a complex network \cite{}. A complex network can be defined is a network that exhibits non-trivial topological properties, which we will explain later on this thesis. These properties are often found in social networks, such as a social media \cite{dunbar-2015}, the collaboration network of scientists \cite{newman-coll-2001,radicchi-2008}, or international conflicts \cite{hafnerburton-2009, diaz2023network}. In particular, the study of information spreading as a dynamical system on networks has allowed to understand how information spreads through a social system and how consensus emerges.

Contagion of information has been a topic of interest for many social scientists. Eearly theoretical frameworks,influenced by psychological and sociological theories, show how individuals in a crowd lose their sense of self and are more susceptible to the ideas and emotions of the crowd \cite{le2023crowd}. Social imitation of behaviors and ideas was proposed as a mechanism for social change, facilitated by close contact and communication among individuals \cite{kanter-1971}. Similarly, peer pressure was also proposed another possible mechanism, where individuals are influenced by their peers to adopt certain behaviors or ideas \cite{granovetter-1978, brown-1986}. However, until now these theoretical dissertations were not supported by quantitative results with real data analysis. For example, the spread of innovations of innovations \cite{rogers2014}, the diffusion of information \cite{valente-1996}, or the spread of diseases \cite{anderson1991infectious} are some examples of social phenomena where we can test if the traditional theoretical frameworks are able to reproduce the data.

Beyond the network structure and the contagion dynamics, there is a third ingredient that is critical for the study of social systems: the human temporal interactions. Human interactions exhibit complex activity patterns that are difficult to understand and to model, since there are a lot of mechanisms that drive the human behavior. For example, the human activity patterns can be driven by the circadian rhythms \cite{roenneberg-2013}, bursty interactions \cite{Barabasi2005Bursts}, cascades \cite{watts-2002}, periodic commuting behaviors \cite{gonzalez2008understanding}, recurring patterns in online behavior \cite{Lazer2009CompSocSci}, etc. All this effects need for a proper characterization. MIRAR LO QUE DIUS MARTON.


\section{\label{sec:Challenges of Computational Social Science} Challenges of Computational Social Science}

- The study of human behavior and social systems is a complex problem that requires the use of computational methods to study human behavior and social systems.

- There are some challenges that are unique to the study of human behavior and social systems, and that are not present in the study of physical systems.

- The main problem is the data availability, and the fact that the data that is generated by human activities is not always available for study.

- Notice that the data sources typically used for the study of human behavior does not come from controlled experiments, but from the digital traces that are generated by human activities.


- The second problem is the data analysis, and the fact that the data that is generated by human activities is not always easy to analyze.

- The data source to analyze usually is a piece of a larger dataset, so we need to be careful to avoid biases in the analysis driven by the data size.

- Temporal windows are also a problem, because when we analyze the dynamics of a system, we need to be careful to avoid biases in the analysis driven by the temporal window.

- The third problem is the modeling, and the fact that the data that is generated by human activities is not always easy to model. 

- Deterministic models are not always useful to model human behavior, and we need to use stochastic models to model human behavior.

- Also, mechanistic models and data driven models is something that we need to consider when we model human behavior.

- Another possibility is to use agent-based models to model human behavior.
With the advent of computational methods in the latter half of the 20th century, researchers gained powerful tools to simulate and analyze complex social systems. Agent-based modeling (ABM) emerged as a particularly influential approach, enabling scientists to create and study systems of interacting agents (individuals or collective entities) and observe emergent behaviors from simple rules of interaction. 

- Computational social science has many applications, and it is being used to study human behavior and social systems.

- Sociotechnical systems, social networks, and human dynamics are some of the applications of computational social science.

- fake news detection, information spreading, and social influence are some of the applications of computational social science.

\section{\label{sec:Terminology and general concepts} Terminology and general concepts}

\subsection{\label{subsec:Complex networks} Complex networks}

As I understood it, what makes a network complex is having a short path length, such that average distance between two nodes is relatively small, high levels of clustering, such that nodes tend to form triangles, and a degree distribution with a fat tail, such that there are a few nodes with a very high degree. These properties are often found in social networks, such as a social media, the collaboration network of scientists, or international conflicts.

- In this section, we introduce some terminology and general concepts that are used in the study of human behavior and social systems.

Complex networks, interface density, and community structure are some of the concepts that are used in the study of human behavior and social systems.

\subsection{\label{subsec:Models} Models}

binary state models, random networks, configuration models, and preferential attachment are some of the models that are used in the study of human behavior and social systems.

\section{\label{sec:Datasets} Datasets}

- We used the idealista dataset

- The strong point of the idealista dataset is that it contains information about the real estate market in Spain, and that it is a large dataset that contains information about the real estate market in Spain.

- The missing point of the idealista dataset is that it contains information about the real estate market in Spain, and that it is a large dataset that contains information about the real estate market in Spain.
