In this concluding chapter, we offer a comprehensive overview of our most significant findings and their implications. We have explored in this thesis how aging and temporal dynamics take a relevant role in shaping the behavior in socio-economic complex systems. Through this thesis, we use a combination of theoretical and empirical approaches to study the emergent dynamics of complex social systems from a quantitative perspective, using tools from statistical physics, network theory, and data science. In particular, we have followed two complementary paths: the first one is the study of aging in threshold models, and the second one is the analysis of the temporal and spatial dynamics of the housing market from real online listings. We have found that aging and memory effects play a crucial role in the dynamics of these systems, leading to new insights and modeling challenges due to the non-Markovian nature of the dynamics.

\section{Remarks on Aging in threshold models \label{sec:aging_threshold_models}}

In the first part of the thesis, we delved into the analysis of aging in threshold models, in particular 3 different models: a segregation model (Sakoda-Schelling), a complex contagion model (Granovetter-Watts), and a consensus model (Symmetrical Threshold model). Aging in this context is introduced as a mechanism that modifies the microdynamics of the system, such that agents become increasingly resistant to change state or location the longer they remain in it. This mechanism is inspired by the empirical observation that individuals tend to stick to their decisions or locations over time, which can be seen as a form of inertia. This non-Markovian mechanism is found to be an important modification for both the stationary state and the dynamics of these models. In particular, in Chapter 3, we examined the aging effects in the Sakoda-Schelling segregation model. The study introduced aging as a factor where satisfied agents become increasingly resistant to changing their location the longer they remain satisfied in it. This modification led to the disappearance of the phase transition between segregated and mixed phases observed in the original model as the tolerance is increased, resulting in a phase diagram with segregated states even for high values of tolerance. This indicates that aging promotes segregation by creating a strong attachment to the current location, which counters the tendency to move based on satisfaction alone. Moreover, the stationary state of the model with aging present a higher global satisfaction than the original model. On the other hand, the coarsening process in the segregated phase was found to be slower in the presence of aging, as the interface density decayed with a lower exponent than in the original model. Furthermore, we found that the aging mechanism is able to break the time translational invariance of the model, leading to a different scaling of the autocorrelation function. For this model, the full study was conducted via numerical simulations, and a an analytical approach should be developed to understand the phase diagram of both the original and the version with aging.

Chapter 4 focused on extending the Approximate Master Equation (AME) for binary-state models to non-Markovian dynamics associated with aging. The original AME was developed to describe the dynamics of binary-state models in complex networks via a set of coupled differential equations where we treat sets of nodes with same degree and same number of neighbors in state $-1$ as compartments of a compartmental model. We extended this approach to include the aging mechanism, where age is treated as an additional variable of a node, as the state or the degree. With this framework, we were able to capture the non-Markovian dynamics associated with aging via two mechanisms: aging, when a node remains at its state and increases age, and resetting, when a node remains at its state and resets its age. With this two simple mechanisms, the AME is able to capture the non-Markovian dynamics for a large variety of binary state models with aging. The most important restriction of the AME is that it assumes an infinite large network with negligible levels of clustering. Nevertheless, this approach could be modified to include network size and stochastic effects related with aging, along the lines of Ref. \cite{peralta-2020B}. In this thesis, Chapter 4 was very useful as a preamble for the analysis of the models presented in the following chapters.

In Chapter 5, we analyzed the impact of aging in the Granovetter-Watts model, a fundamental model for understanding complex contagion processes. In this context, we introduced two aging mechanisms: exogenous aging, where agents become increasingly resistant to adopting a behavior the longer they remain inactive, and endogenous aging, where agents become increasingly resistant to adopting a behavior the longer they remain in its state, even if the agents activate. For both mechanisms, aging was shown to slow down the cascade dynamics without altering the cascade condition. The model's behavior transitioned from exponential to stretched exponential or power-law adoption dynamics, depending on the aging mechanism (exogenous and endogenous). This chapter provided analytical expressions for the cascade conditions and exponents, derived from the AME, offering a comprehensive understanding of how aging influences the spread of information or behaviors in social networks. Furthermore, the results of this model in a Moore lattice were analyzed via numerical simulations, as the AME formalism is not valid for this network structure. In this regular network, the aging mechanism slows down the cascade dynamics dramatically, leading to very slow increasing adoption curves.

Chapter 6 was divided into two parts, addressing the phase diagram and the dynamics of the Symmetrical Threshold model, a consensus threshold model where both states are symmetric. The first part (Chapter 6A) discussed the ordering dynamics without aging, presenting a detailed analysis of the phase diagram and identifying three distinct phases: mixed, ordered, and frozen. The dynamical regimes of the model were characterized by a set of observables that allowed to distinguish the different phases, which were found to be in good agreement with the theoretical predictions of the AME. The second part (Chapter 6B) explored the implications of endogenous aging in this model. For sparse network, aging introduced a new dynamical phase, at the mixed region of the phase diagram, characterized by an initial disordering followed by slow ordering. In this phase, it was found that consensus can be reached in the state of an initial minority due to aging. Moreover, aging significantly altered the interface density decay and persistence, showing that aging leads to a slower convergence towards the ordered state. The AME derived in Chapter 4 was used to analyze the model, providing a comprehensive understanding of the aging effects on consensus dynamics driven by group interactions. For both, the aging and non-aging models, the results were shown for 4 different network topologies (Complete graph, Erd\H{o}s-Rényi, random-regular and Moore lattice), showing how aging implications change for different network structures.

\section{Remarks on Assessing the housing market dynamics \label{sec:housing_market_dynamics}}

In the second part of the thesis, keeping this idea of burstiness and memory effects, we analyzed the static and dynamical properties of a real complex system: the housing market. We found that there are irregular temporal patterns in the listings data, which affect the strategic interactions between sellers and real estate agencies. Moreover, we identified that there is a spatial segmentation in the housing market, driven by the agencies influence and specialization.

Motivated by the spatial specialization of the real estate agencies, Chapter 7 explores the spatial segmentation of the housing market from online listings in 3 provinces in Spain and France. By developing a novel methodology to evaluate spatial segmentation using listings data, we identified robust submarkets defined by the influence of real estate agencies. This segmentation was found to be consistent across different spatial resolutions and community detection algorithms. When applied to census level data, we found that the identified communities differ from the ones using cells or municipalitiy level data. To avoid this particular scale-depencency of the detected submarkets, we proposed a stochastically aggregative method to obtain a coherent segmentation from the data at this level. This methodology was proven to be useful and allowed to identify communities from the French dataset, where the identified communities showed similar features with ones obtained from the Spanish dataset (submarkets larger than municipalites, with connected spatial structure). This suggests that the submarket partition is a result of an underlying robust mechanism and that market-based spatial divisions are crucial for effective policy-making and economical price modeling.

In Chapter 8, we extended the previous analysis to study the dynamics of real estate listings in 3 Spanish provinces, focusing on the listings temporal dynamics, the decision-making processes of house sellers and the role of real estate agencies. The analysis revealed that the temporal patterns in the listings show a persistence effect similar to aging, where the longer a listing remains available on the market, the less likely is to be sold. This effect was found to be consistent across different provinces and property types analyzed, suggesting that it is a general feature of the housing market. Additionally, via a temporal bipartite network analysis, we were able to identify via a quantitative approach the key factors influencing the decision-making process of house sellers when choosing a real estate agency. The results showed that there is a preferential attachment mechanism in the selection of agencies (where the most active agencies tend to attract more listings), a price similarity effect (where listings are posted by agencies with similar prices), and spatial specialization (where agencies tend to focus on specific areas). These findings provide a comprehensive understanding of the strategic interactions between sellers and agencies in the housing market from a quantitative perspective, with agreement with qualitative observations in the literature.

In conclusion, the findings of this thesis underscore the profound impact of aging and temporal dynamics on socio-economic systems. By incorporating aging into various models, we have demonstrated how memory effects alter system behaviors, leading to new insights into segregation dynamics, contagion processes, and consensus problems. Moreover, the analysis of real-world data from the housing market has revealed the importance of temporal (memory) and spatial (specialization) dynamics in shaping market structures and decision-making processes. These results not only advance theoretical understanding but also provide practical implications for policy and strategy in social and economic contexts.
