\section{\label{sec:Introduction} Introduction}

- Modeling human dynamics in opinion formation processes, or technology adoption process has been object of study for many years. 

- The contagion of ideas is a process that has been studied for many years, and it is a process that is present in many social systems.

binary-state models are a versatile tool to describe a variety of natural and social phenomena in systems formed by many interacting agents. Each agent is considered to be in one of two possible states: susceptible/infected, adopters/non-adopters, democrat/republican, etc, depending on the context of the model. The interaction among agents is determined by the underlying network and the dynamical rules of the model. There are many examples of binary-state models, including processes of opinion formation \cite{Voter-original,sood-2005,fernandez-gracia-2014,redner-2019}, disease or social contagion \cite{granovetter-1978,pastor-satorras-2015}, etc. Extended and modified versions of these models can lead to very different dynamical behaviors than in the original model. As examples, the use of multi-layer  \cite{diakonova-2014,diakonova-2016,amato-2017} or time-dependent networks \cite{vazquez-2008}, higher-order interactions \cite{de-arruda-2020, iacopini-2019, cencetti-2021}, non-linear collective phenomena \cite{castellano-2009,peralta-2018}, noise \cite{carro-2016} and non-Markovian \cite{van-mieghem-2013,starnini-2017,peralta-2020A,chen-2020} effects induce significant changes to the dynamics.

\dots

\section{\label{sec:Simple and Complex Contagion} Simple and Complex Contagion}

- The contagion of ideas is a process that has been studied for many years, and it is a process that is present in many social systems. 

- Simple contagion is used to describe the process of contagion of ideas that is driven by the number of contacts that an individual has with other individuals.

- Complex contagion is used to describe the process of contagion of ideas that is driven by the number of contacts that an individual has with other individuals, and by the number of contacts that the individuals that the individual has contact with have with other individuals.

\section{\label{sec:Threshold model} Threshold model}

- The threshold model is used to describe processes of complex contagion based on the idea that individuals have a threshold that needs to be reached in order to adopt an idea.

- Watts (2002) proposed a simple model to study the dynamics of complex contagion based on the idea that individuals have a threshold that needs to be reached in order to adopt an idea. 

- Should I describe what is a phase transition? Maybe with a foot note.

- This model exhibits a phase transition from a regime where the adoption of an idea is rare to a regime where the adoption of an idea is common. This phase transition is discontinuous.

- This model has been studied for many social networks and many variants of the model have been proposed\dots

\section{\label{The Sakoda-Schelling model} The Sakoda-Schelling model}

- Thomas C. Schelling (1975) proposed a model to study the dynamics of segregation based on the idea that individuals have a threshold that needs to be reached in order to move to a different location. 

- This model become very popular because it was able to reproduce the segregation patterns that are observed in many cities, just as an emergent phenomena from individual decisions.

- On the other hand, this model was already studied by Sakoda in 1971, in a paper that was published in Japanese, and that was not known by the scientific community. In this work, the author proposed a model to study the dynamics of segregation based on the idea that individuals have a threshold that needs to be reached in order to move to a different location.

- Schelling model is a particular case of the Sakoda model.

- This model exhibits a phase transition from a regime where the segregation is low to a regime where the segregation is high. This phase transition is discontinuous. 

Thomas Schelling introduced a simple segregation model   \cite{schelling-1969,Schelling,schellingbook,hegselmann-2017} in which agents of two colors are distributed randomly on a chess-board, leaving some locations free. Agents are unsatisfied if more than a half of the eight nearest neighbors have different color. Randomly, the unsatisfied agents will move to available satisfying locations of the neighborhood. This model has had a very significant impact for several reasons: The ``hand-made'' simulations performed by T. Schelling by moving pawns on a chessboard are an early precedent of the use of  agent-based simulations in Social Sciences. It is also one of the first social models to show emergent behavior as a result of simple interactions among agents, a characteristic of complex systems.

- Despite the many variants of the model that have been proposed, the phase transition emergent in this model is very robust.

\section{\label{sec: Bursty Human Dynamics} Bursty Human Dynamics}

- All models previously described are based in an important assumption regarding to its dynamics: the interactions between individuals occur at a constant rate.

- This constant rate assumption is assuming that the stochastic interactions follow a Poisson process. 

- However, from the analysis of many datasets of human interactions, it has been observed that the interactions between individuals are not constant, but bursty.

- The bursty behaviour is characterized by power law interevent time distributions, and it is present in many human activities, such as e-mail communication, face-to-face interactions, and phone calls.

- The modeling of bursty human dynamics is important because it allows us to understand how the bursty behaviour of human interactions affects the dynamics of social systems.

- Previous literature has several approaches to model bursty human dynamics: temporal networks, activity-driven models, aging \dots

\section{\label{sec:Aging mechanism} Aging mechanism}

- The aging mechanism is a mechanism that has been proposed to model the bursty behaviour of human interactions.

- Aging mechanism is based on the idea that the probability of an individual to interact with another individual decreases with the time since the last interaction.

- Attachment to previous beliefs or habits is a common feature in human behavior. Granovetter (1973) discussed 

- Instead of the constant rate assumption, the aging mechanism assumes that the interactions between individuals occur at a rate that decreases with the time since the last interaction.

- The focus on this approach is to include the bursty dynamics in the individuals attempts to interact with others, rather than in the interactions themselves (as in the activity-driven models or temporal networks).

It is known that human interactions do not occur at a constant rate. They rather show  a bursty character with a non-Poissonian inter-event time distribution that reflects a memory from past interactions. \cite{barabasi-2005,moro,oriol,rybski-2012,zignani-2016,kumar-2020}
However, most social simulations, including simulations of variants of the Sakoda-Schelling model, implicitly assume a constant rate of interactions or state updating. "Aging" is one form of memory effect on which the rate of interactions depends on the persistence time of an agent in a state, modifying the transition to a different state \cite{fernandez-gracia-2011,perez-2016,boguna-2014}. This concept of aging, or "social inertia" \cite{Stark2008}, constrains the transitions in a way that the longer an agent remains in a given state, the smaller the probability to change it. Aging has been already shown to modify social dynamics very significantly. For example, in opinion dynamics, aging is able to produce coarsening towards a consensus state in the voter model \cite{fernandez-gracia-2011,peralta-2020} or to induce a continuous phase transition in the noisy voter model \cite{artime-2018}.

\section{\label{sec:Aging in Simple Contagion Models} Aging in Simple Contagion Models}

- Aging in the Voter model has been studied by many authors, and it has been shown that the aging mechanism affects the dynamics of the model.

- The aging mechanism has been shown to affect the dynamics of the Voter model, and to change the phase transition of the model.

- Aging in the noisy voter model is able to change the phase transition of the model, and to make the phase transition continuous.

- Aging in the SI model is able to change the cascade size distribution of the model, and to make the cascade size distribution follow a power law (see the work of Karsai et al. 2011).