\pagebreak
\frontmatter
\thispagestyle{empty}
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgements}
\textbf{ \huge Acknowledgements}

\vspace*{0.5 cm}

First, I would like to express my gratitude to my two directors and main collaborators through this thesis, Jos\'e J. Ramasco and Maxi San Miguel. I am very grateful for their educational and professional guidance through this path, which has been fundamental to reach this point. In particular, Jose has been a constant guide, and I am very grateful that his door always remained open for me, available for any question or doubt that could take 5 minutes or 5 hours to solve. Maxi has been a great supervisor, I want to thank him for his critical view, his constant push to improve my work, and his questions that always made me think in a different way. I must also thank them for their financial support and generosity, which allowed me to do this three-year PhD without worrying much about money. I would also like to thank Vincenzo Nicosia, with whom I had the pleasure to do a research stay during the last year of my PhD. Collaborating with him allowed me to be part of the process of developing new ideas and projects, which has motivated me with new challenges and perspectives. In addition, throughout this period at IFISC, I thank having great conversations with scientists such as Juan Fern\'andez-Gracia, Konstantin Klemm and Damià Gomila, which have served as a source of inspiration and learning.

My firsts steps began in the Master in Physics of Complex Systems, where I had the opportunity to meet great people and friends. I would like to thank my classmates, Alex, Miguel, Marius, Robert, Javi, Medi, Joje, Ana. I want to acknowledge the great environment we had during those years, which motivated me to follow this path. Alex has not only been my master's colleague, I consider him as a partner though this journey. I thank all the moments helping each other, being victims of FPI-CAIB's injustices, promoting Julia as the best programming language and doing ``rutillas chill'' to both Massanella and Sabotage. I am very grateful that we did this path together, and I hope to be able to work someday in the Alex's Research Institute. Also, I want to thank Miguel for all the random videos during the pandemic, the curiosities in Google Maps and the funny random videos.

I also want to thank the really nice environment between PhD students and PostDocs at IFISC. It may seem something small, but you just need to go to work once alone to realize how important it is. From all IFISC Basement, I need to highlight a special place for me, the ``Zulo'', which was an almost-empty temporary lab of Nonlinear Physics when I got there, and we turned into a cozy place to work and fight for justice via posters. I want to thank being part of this group and having worked in (probably) the only theoretical physics office with a giant extractor hood.

Durante el doctorado, estoy agradecido de haber conocido a gente maravillosa como Manu. Agradezco las muchas (muchas muchas) conversaciones que hemos tenido a lo largo de estos años, ya sean a nivel profesional como personal, ayudándonos con nuestras epidemias raras. Recordaré las vueltas a casa quedándonos en la esquina charlando incluso sobre lo mucho que charlamos. Has sido de verdad mi Recursos HuManus y con mucho cariño te digo que parte de esta tesis también es tuya. También quiero agradecer haber conocido a Mar, por todo el apoyo, por todas las risas y por animarme a ser más fuerte. Sabes que yo siempre seré el primer MAG y tú mi MAH, ya que con nadie más tienes un póster de ensaimadas. También agradecer a Gorka, por las risas, por las conversaciones sobre porque ningún método de clustering va bien y por mantenerme al día sobre internet.com. Por supuesto, Fer ha sido un gran apoyo durante este periodo, agradezco las largas discusiones en el piso sobre mil temas, creo que no hay proyecto en el IFISC que no hayamos debatido y pensado como lo haríamos nosotros. Gracias por todas las referencias que me has dado y por demostrarme empíricamente que uno puede dormirse donde quiera. Cuando llegasteis al doctorado estaba un poco perdido y os agradezco haberme ayudado a encontrar mi camino.

També vull agrair a l'Adrià, per haver compartit moments graciosos, pel Philomeno Moreno, per l'ADSL i per les millors versions d'"Un Mundo Ideal". También estoy muy agradecido por haber compartido momentos con Bea (gracias por los memes de patitos), Pau, Maria, Mar F., Irene, Lisa, Rodri, Jose. Estoy agradecido de haber podido hacer el doctorado con vosotros, pero sobre todo por todos los momentos que no eran doctorado. Una mención especial a mi equipo de la "Grieta del Invocador": Jesus, Jun, Manu, Giovani. Ojalá saquemos un paper de esto eventualmente, pero si no lo hacemos, nos habremos llevado unas risas. Y por último, a mis compañeros del Barco, Alba y Manu P., que aunque estuvimos a punto de hundirnos varias veces, fue un placer navegar con vosotros

Vull també agrair als meus amics de tota la vida: Jaume, Vernet, Alberto i Marcel. Us agraeixo haver estat allà sempre per desfogar-me de la feina i per donar-me un punt de vista diferent. En aquesta feina, a vegades et perds una mica de la realitat i m'heu ajudat a tocar de peus a terra. També un agraïment molt fort i sincer a la Rosa i el Pau C., per tot el suport permetent-me parlar de qualsevol tema sense filtre i sentir-me escoltat i entès. Us agraeixo molt haver estat allà per mi i haver-me ajudat a superar moments difícils.

I per descomptat, a la meva família, al meu pare, a la meva mare, a les meves germanes Adriana i Ivet, al meu fillol Aleix, que és més jove que aquesta tesi, a la Jose, al Pau, a la Mimi. Gràcies per tot el suport i per la paciència quan us parlo d'alguna cosa que no enteneu. De debò que no treballo amb ocells però és un bon exemple de sistema complex. Gràcies també per tot el suport tant financer com emocional durant tots els anys que m'han dut a aquest punt. Òbviament, sense valtros no hi hauria tesi.

Finalment, vull agrair a una persona en especial que ha estat al meu costat des de l'inici d'aquesta tesi i amb la que hem construït junts un projecte de vida en comú. Riki, gràcies per ser la pedra angular que ha aguantat aquest procés, ho dic de debò quan dic que sense tu, aquesta tesi no existiria. Estic molt agraït per tots els missatges d'àudio llargs que t'ha tocat escoltar, per totes les converses sobre compaginar feina i vida, per donar-me suport en moments de frustració i entendre que a vegades necessitava estar sol, i per intentar trobar solució a problemes que ni tan sols jo sabia per on començar. Per tot això i per molt més, mereixes més que un simple agraïment. Aquesta tesi és tan teva com meva.


\thispagestyle{empty}


\vfill